### 2022-06-19 draft.tex
Deploying to gh-pages from @ PoisotLab/ms_metaweb_perspectives@7d1f33e481bf1cda2e0b88f5b79db035106f84c9 🚀

One prominent family of approaches we do not discuss in the present
manuscript is Graph Neural Networks (GNN; \textbf{Zhou2020GraNeu?}); GNN
are, in a sense, a method to embed a graph into a dense subspace, but
belong to the family of deep learning methods, which has its own set of
practices (see \emph{e.g.} \textbf{Goodfellow2016DeeLea?}). An important
issue with methods based on deep learning is that because their
parameter space is immense, the sample size of the data fed into them
must be similarly large (typically thousands of instances). This is a
requirement for the model to converge correctly during training, but
this assumption is unlikely to be met given the size of datasets about
currently available metawebs (or single time/location species
interaction networks). This data volume requirement is mostly absent
from the techniques we list below. Furthermore, GNN still have some
challenges related to their shallow structure, and concerns related to
scalability (see \textbf{Gupta2021GraNeu?} for a review), which are
mostly absent from the methods listed in tbl.~\ref{tbl:methods}.
Assuming that the uptake of next-generation biomonitoring techniques
does indeed deliver larger datasets on species interactions
(\textbf{Bohan2017NexGlo?}), there is a potential for GNN to become an
applicable embedding/predictive technique in the coming years.

@@

Graph embeddings \emph{can} serve as a dimensionality reduction method.
For example, RDPG (Strydom, Bouskila, et al., 2021) and t-SVD (Poisot et
al., 2021) typically embed networks using fewer dimensions than the
original network (the original network has as many dimensions as
species, and as many informative dimensions as trophically unique
species; Strydom, Dalla Riva, et al., 2021). But this is not
necessarilly the case -- indeed, one may perform a PCA (a special case
of SVD) to project the raw data into a subspace that improves the
efficacy of t-SNE (\textbf{Maaten2009LeaPar?}). There are many
dimensionality reductions (\textbf{Anowar2021ConEmp?}) that can be
applied to an embedded network should the need for dimensionality
reduction (for example for data visualisation) arise. In brief, many
graph embeddings \emph{can} serve as dimensionality reduction steps, but
not all do, neither do all dimensionality reduction methods provide
adequate graph embedding capacities.

@@

In this section, we illustrate the embedding of a collection of
bipartite networks collected by \textbf{REF}, using truncated Singular
Value Decomposition (t-SVD). The code to reproduce this example (as a
Jupyter notebook) is available as supplementary material. The resulting
(binary) metaweb \(\mathcal{M}\) has 2131 interactions between 206
parasites and 121 hosts (the same metaweb was used for the illustration
of interaction inference through deep learning in Strydom, Catchen, et
al., 2021), and its adjacency matrix has full rank (\emph{i.e.} it
represents a space with 121 dimensions). All analyses were done using
Julia (Bezanson et al., 2017) version 1.7.2, \emph{Makie.jl}
(\textbf{Danisch2021MakJl?}), and \emph{EcologicalNetworks.jl} (Poisot
et al., 2019).

@@

The final family of problems relates less to ecological methods than to
the praxis of ecological research. Operating under the context of
national divisions, in large parts of the world, reflects nothing more
than the legacy of settler colonialism, which drives a disparity in
available ecological data. Applying any embedding to biased data does
not debias them, but instead embeds these very same biases, propagating
them to the machine learning models using embeddings tomake predictions.
Indeed, the use of ecological data is not an apolitical act (Nost \&
Goldstein, 2021), as data infrastructures tend to be designed to answer
questions within national boundaries (therefore placing contingencies on
what is available to be embedded), and their use often draws upon and
reinforces territorial statecraft. As per Machen \& Nost (2021), this is
particularly true when the output of ``algorithmic thinking''
(\emph{e.g.} relying on machine learning to generate knowledge) can be
re-used for governance (\emph{e.g.} enacting conservation decisions at
the national scale). As information on species interaction networks
structure is increasingly leveraged as a tool to guide conservation
actions (see \emph{e.g.} recent discussions for food-web based
conservation; \textbf{Eero2021UseFoo?}; \textbf{NamanFooWeb?};
\textbf{Stier2017IntExp?}), the need to appraise and correct biases that
are unwittingly propagated to algorithms when embedded from the original
data is paramount. Predictive approaches deployed at the continental
scale, no matter their intent, originate in the framework that
contributed to the ongoing biodiversity crisis (Adam, 2014) and
reinforced environmental injustice (Choudry, 2013; Domínguez \& Luoma,
2020). Particularly on Turtle Island and other territories that were
traditionally stewarded by Indigenous people, these approaches should be
replaced (or at least guided and framed) by Indigenous principles of
land management (Eichhorn et al., 2019; No'kmaq et al., 2021), as part
of an ``algorithm-in-the-loop'' approach. Human-algorithm interactions
are notoriously difficult and can yield adverse effect
(\textbf{Green2019DisInt?}; \textbf{Stevenson2021AlgRis?}), suggesting
the need to systematically study them for the specific purpose of
biodiversity governance, as well as to improve the algorithmic literacy
of decision makers. As we see artificial intelligence/machine learning
being increasingly mobilized to generate knowledge that is lacking for
conservation decisions (\emph{e.g.} Lamba et al., 2019; Mosebo Fernandes
et al., 2020) and drive policy decisions (Weiskopf et al., 2022), our
discussion of these tools need to go beyond the technical and
statistical, and into the governance consequences they can have.

@@

\leavevmode\hypertarget{ref-Arsov2019NetEmb}{}%
Arsov, N., \& Mirceva, G. (2019, November 26). \emph{Network Embedding:
An Overview}. \url{http://arxiv.org/abs/1911.11726}

\leavevmode\hypertarget{ref-Bartomeus2016ComFra}{}%
Bartomeus, I., Gravel, D., Tylianakis, J. M., Aizen, M. A., Dickie, I.
A., \& Bernard-Verdier, M. (2016). A common framework for identifying
linkage rules across different types of interactions. \emph{Functional
Ecology}, \emph{30}(12), 1894--1903.

@@

@@

\leavevmode\hypertarget{ref-Chen2017DeeMul}{}%
Chen, D., Xue, Y., Fink, D., Chen, S., \& Gomes, C. P. (2017).
\emph{Deep Multi-species Embedding}. 3639--3646.

\leavevmode\hypertarget{ref-Chen2017HarHie}{}%
Chen, H., Perozzi, B., Hu, Y., \& Skiena, S. (2017, November 16).
\emph{HARP: Hierarchical Representation Learning for Networks}.
\url{http://arxiv.org/abs/1706.07845}

@@

@@

@@

\leavevmode\hypertarget{ref-Fricke2022EffDef}{}%
Fricke, E. C., Ordonez, A., Rogers, H. S., \& Svenning, J.-C. (2022).
The effects of defaunation on plants' capacity to track climate change.
\emph{Science}.

@@

\leavevmode\hypertarget{ref-Galiana2019GeoVar}{}%
Galiana, N., Hawkins, B. A., \& Montoya, J. M. (2019). The geographical
variation of network structure is scale dependent: Understanding the
biotic specialization of hostparasitoid networks. \emph{Ecography},
\emph{42}(6), 1175--1187. \url{https://doi.org/10.1111/ecog.03684}

@@

@@

@@

@@

@@

\leavevmode\hypertarget{ref-Murphy2022ProMac}{}%
Murphy, K. P. (2022). \emph{Probabilistic machine learning: An
introduction}. MIT Press.

\leavevmode\hypertarget{ref-Narayanan2017GraLea}{}%
Narayanan, A., Chandramohan, M., Venkatesan, R., Chen, L., Liu, Y., \&
Jaiswal, S. (2017, July 17). \emph{Graph2vec: Learning Distributed
Representations of Graphs}. \url{http://arxiv.org/abs/1707.05005}

@@

\leavevmode\hypertarget{ref-OConnor2020UnvFoo}{}%
O'Connor, L. M. J., Pollock, L. J., Braga, J., Ficetola, G. F.,
Maiorano, L., Martinez-Almoyna, C., Montemaggiori, A., Ohlmann, M., \&
Thuiller, W. (2020). Unveiling the food webs of tetrapods across Europe
through the prism of the Eltonian niche. \emph{Journal of Biogeography},
\emph{47}(1), 181--192. \url{https://doi.org/10.1111/jbi.13773}

@@

\leavevmode\hypertarget{ref-Poisot2021ImpMam}{}%
Poisot, T., Ouellet, M.-A., Mollentze, N., Farrell, M. J., Becker, D.
J., Albery, G. F., Gibb, R. J., Seifert, S. N., \& Carlson, C. J. (2021,
May 31). \emph{Imputing the mammalian virome with linear filtering and
singular value decomposition}. \url{http://arxiv.org/abs/2105.14973}

@@

\leavevmode\hypertarget{ref-Ramasamy2015ComSpe}{}%
Ramasamy, D., \& Madhow, U. (2015). Compressive spectral embedding:
Sidestepping the SVD. In C. Cortes, N. Lawrence, D. Lee, M. Sugiyama, \&
R. Garnett (Eds.), \emph{Advances in neural information processing
systems} (Vol. 28). Curran Associates, Inc.

@@

@@

\leavevmode\hypertarget{ref-Xu2020UndGra}{}%
Xu, M. (2020, December 14). \emph{Understanding graph embedding methods
and their applications}. \url{http://arxiv.org/abs/2012.08019}

@@

>>>>

One prominent family of approaches we do not discuss in the present
manuscript is Graph Neural Networks (GNN; Zhou et al., 2020); GNN are,
in a sense, a method to embed a graph into a dense subspace, but belong
to the family of deep learning methods, which has its own set of
practices (see \emph{e.g.} Goodfellow et al., 2016). An important issue
with methods based on deep learning is that because their parameter
space is immense, the sample size of the data fed into them must be
similarly large (typically thousands of instances). This is a
requirement for the model to converge correctly during training, but
this assumption is unlikely to be met given the size of datasets about
currently available metawebs (or single time/location species
interaction networks). This data volume requirement is mostly absent
from the techniques we list below. Furthermore, GNN still have some
challenges related to their shallow structure, and concerns related to
scalability (see Gupta et al., 2021 for a review), which are mostly
absent from the methods listed in tbl.~\ref{tbl:methods}. Assuming that
the uptake of next-generation biomonitoring techniques does indeed
deliver larger datasets on species interactions (Bohan et al., 2017),
there is a potential for GNN to become an applicable
embedding/predictive technique in the coming years.

@@

Graph embeddings \emph{can} serve as a dimensionality reduction method.
For example, RDPG (Strydom, Bouskila, et al., 2021) and t-SVD (Poisot et
al., 2021) typically embed networks using fewer dimensions than the
original network (the original network has as many dimensions as
species, and as many informative dimensions as trophically unique
species; Strydom, Dalla Riva, et al., 2021). But this is not
necessarilly the case -- indeed, one may perform a PCA (a special case
of SVD) to project the raw data into a subspace that improves the
efficacy of t-SNE (Maaten, 2009). There are many dimensionality
reductions (Anowar et al., 2021) that can be applied to an embedded
network should the need for dimensionality reduction (for example for
data visualisation) arise. In brief, many graph embeddings \emph{can}
serve as dimensionality reduction steps, but not all do, neither do all
dimensionality reduction methods provide adequate graph embedding
capacities.

@@

In this section, we illustrate the embedding of a collection of
bipartite networks collected by \textbf{REF}, using truncated Singular
Value Decomposition (t-SVD). The code to reproduce this example (as a
Jupyter notebook) is available as supplementary material. The resulting
(binary) metaweb \(\mathcal{M}\) has 2131 interactions between 206
parasites and 121 hosts (the same metaweb was used for the illustration
of interaction inference through deep learning in Strydom, Catchen, et
al., 2021), and its adjacency matrix has full rank (\emph{i.e.} it
represents a space with 121 dimensions). All analyses were done using
Julia (Bezanson et al., 2017) version 1.7.2, \emph{Makie.jl} (Danisch \&
Krumbiegel, 2021), and \emph{EcologicalNetworks.jl} (Poisot et al.,
2019).

@@

The final family of problems relates less to ecological methods than to
the praxis of ecological research. Operating under the context of
national divisions, in large parts of the world, reflects nothing more
than the legacy of settler colonialism, which drives a disparity in
available ecological data. Applying any embedding to biased data does
not debias them, but instead embeds these very same biases, propagating
them to the machine learning models using embeddings tomake predictions.
Indeed, the use of ecological data is not an apolitical act (Nost \&
Goldstein, 2021), as data infrastructures tend to be designed to answer
questions within national boundaries (therefore placing contingencies on
what is available to be embedded), and their use often draws upon and
reinforces territorial statecraft. As per Machen \& Nost (2021), this is
particularly true when the output of ``algorithmic thinking''
(\emph{e.g.} relying on machine learning to generate knowledge) can be
re-used for governance (\emph{e.g.} enacting conservation decisions at
the national scale). As information on species interaction networks
structure is increasingly leveraged as a tool to guide conservation
actions (see \emph{e.g.} recent discussions for food-web based
conservation; Eero et al., 2021; Naman et al., 2022; Stier et al.,
2017), the need to appraise and correct biases that are unwittingly
propagated to algorithms when embedded from the original data is
paramount. Predictive approaches deployed at the continental scale, no
matter their intent, originate in the framework that contributed to the
ongoing biodiversity crisis (Adam, 2014) and reinforced environmental
injustice (Choudry, 2013; Domínguez \& Luoma, 2020). Particularly on
Turtle Island and other territories that were traditionally stewarded by
Indigenous people, these approaches should be replaced (or at least
guided and framed) by Indigenous principles of land management (Eichhorn
et al., 2019; No'kmaq et al., 2021), as part of an
``algorithm-in-the-loop'' approach. Human-algorithm interactions are
notoriously difficult and can yield adverse effect (Green \& Chen, 2019;
Stevenson \& Doleac, 2021), suggesting the need to systematically study
them for the specific purpose of biodiversity governance, as well as to
improve the algorithmic literacy of decision makers. As we see
artificial intelligence/machine learning being increasingly mobilized to
generate knowledge that is lacking for conservation decisions
(\emph{e.g.} Lamba et al., 2019; Mosebo Fernandes et al., 2020) and
drive policy decisions (Weiskopf et al., 2022), our discussion of these
tools need to go beyond the technical and statistical, and into the
governance consequences they can have.

@@

\leavevmode\hypertarget{ref-Anowar2021ConEmp}{}%
Anowar, F., Sadaoui, S., \& Selim, B. (2021). Conceptual and empirical
comparison of dimensionality reduction algorithms (PCA, KPCA, LDA, MDS,
SVD, LLE, ISOMAP, LE, ICA, t-SNE). \emph{Computer Science Review},
\emph{40}, 100378. \url{https://doi.org/10.1016/j.cosrev.2021.100378}

\leavevmode\hypertarget{ref-Arsov2019NetEmb}{}%
Arsov, N., \& Mirceva, G. (2019). \emph{Network Embedding: An Overview}.
\url{http://arxiv.org/abs/1911.11726}

\leavevmode\hypertarget{ref-Bartomeus2016ComFra}{}%
Bartomeus, I., Gravel, D., Tylianakis, J. M., Aizen, M. A., Dickie, I.
A., \& Bernard-Verdier, M. (2016). A common framework for identifying
linkage rules across different types of interactions. \emph{Functional
Ecology}, \emph{30}(12), 1894--1903.
\url{http://onlinelibrary.wiley.com/doi/10.1111/1365-2435.12666/full}

@@

\leavevmode\hypertarget{ref-Bohan2017NexGlo}{}%
Bohan, D. A., Vacher, C., Tamaddoni-Nezhad, A., Raybould, A., Dumbrell,
A. J., \& Woodward, G. (2017). Next-Generation Global Biomonitoring:
Large-scale, Automated Reconstruction of Ecological Networks.
\emph{Trends in Ecology \& Evolution}.
\url{https://doi.org/10.1016/j.tree.2017.03.001}

@@

\leavevmode\hypertarget{ref-Chen2017DeeMul}{}%
Chen, D., Xue, Y., Fink, D., Chen, S., \& Gomes, C. P. (2017).
\emph{Deep Multi-species Embedding}. 3639--3646.
\url{https://www.ijcai.org/proceedings/2017/509}

\leavevmode\hypertarget{ref-Chen2017HarHie}{}%
Chen, H., Perozzi, B., Hu, Y., \& Skiena, S. (2017). \emph{HARP:
Hierarchical Representation Learning for Networks}.
\url{http://arxiv.org/abs/1706.07845}

@@

\leavevmode\hypertarget{ref-Danisch2021MakJl}{}%
Danisch, S., \& Krumbiegel, J. (2021). Makie.jl: Flexible
high-performance data visualization for Julia. \emph{Journal of Open
Source Software}, \emph{6}(65), 3349.
\url{https://doi.org/10.21105/joss.03349}

@@

\leavevmode\hypertarget{ref-Eero2021UseFoo}{}%
Eero, M., Dierking, J., Humborg, C., Undeman, E., MacKenzie, B. R.,
Ojaveer, H., Salo, T., \& Köster, F. W. (2021). Use of food web
knowledge in environmental conservation and management of living
resources in the Baltic Sea. \emph{ICES Journal of Marine Science},
\emph{78}(8), 2645--2663. \url{https://doi.org/10.1093/icesjms/fsab145}

@@

\leavevmode\hypertarget{ref-Fricke2022EffDef}{}%
Fricke, E. C., Ordonez, A., Rogers, H. S., \& Svenning, J.-C. (2022).
The effects of defaunation on plants' capacity to track climate change.
\emph{Science}.
\url{https://www.science.org/doi/abs/10.1126/science.abk3510}

@@

\leavevmode\hypertarget{ref-Galiana2019GeoVar}{}%
Galiana, N., Hawkins, B. A., \& Montoya, J. M. (2019). The geographical
variation of network structure is scale dependent: Understanding the
biotic specialization of host--parasitoid networks. \emph{Ecography},
\emph{42}(6), 1175--1187. \url{https://doi.org/10.1111/ecog.03684}

@@

\leavevmode\hypertarget{ref-Goodfellow2016DeeLea}{}%
Goodfellow, I., Bengio, Y., \& Courville, A. (2016). \emph{Deep
learning}. MIT Press.

@@

\leavevmode\hypertarget{ref-Green2019DisInt}{}%
Green, B., \& Chen, Y. (2019). Disparate Interactions: An
Algorithm-in-the-Loop Analysis of Fairness in Risk Assessments.
\emph{Proceedings of the Conference on Fairness, Accountability, and
Transparency}, 90--99. \url{https://doi.org/10.1145/3287560.3287563}

@@

\leavevmode\hypertarget{ref-Gupta2021GraNeu}{}%
Gupta, A., Matta, P., \& Pant, B. (2021). Graph neural network: Current
state of Art, challenges and applications. \emph{Materials Today:
Proceedings}, \emph{46}, 10927--10932.
\url{https://doi.org/10.1016/j.matpr.2021.01.950}

@@

\leavevmode\hypertarget{ref-Maaten2009LeaPar}{}%
Maaten, L. van der. (2009). Learning a Parametric Embedding by
Preserving Local Structure. \emph{Proceedings of the Twelth
International Conference on Artificial Intelligence and Statistics},
384--391. \url{https://proceedings.mlr.press/v5/maaten09a.html}

@@

\leavevmode\hypertarget{ref-Murphy2022ProMac}{}%
Murphy, K. P. (2022). \emph{Probabilistic machine learning: An
introduction}. MIT Press. \href{https://probml.ai}{probml.ai}

\leavevmode\hypertarget{ref-Naman2022FooWeb}{}%
Naman, S. M., White, S. M., Bellmore, J. R., McHugh, P. A., Kaylor, M.
J., Baxter, C. V., Danehy, R. J., Naiman, R. J., \& Puls, A. L. (2022).
Food web perspectives and methods for riverine fish conservation.
\emph{WIREs Water}, \emph{n/a}(n/a), e1590.
\url{https://doi.org/10.1002/wat2.1590}

\leavevmode\hypertarget{ref-Narayanan2017GraLea}{}%
Narayanan, A., Chandramohan, M., Venkatesan, R., Chen, L., Liu, Y., \&
Jaiswal, S. (2017). \emph{Graph2vec: Learning Distributed
Representations of Graphs}. \url{http://arxiv.org/abs/1707.05005}

@@

\leavevmode\hypertarget{ref-OConnor2020UnvFoo}{}%
O'Connor, L. M. J., Pollock, L. J., Braga, J., Ficetola, G. F.,
Maiorano, L., Martinez‐Almoyna, C., Montemaggiori, A., Ohlmann, M., \&
Thuiller, W. (2020). Unveiling the food webs of tetrapods across Europe
through the prism of the Eltonian niche. \emph{Journal of Biogeography},
\emph{47}(1), 181--192. \url{https://doi.org/10.1111/jbi.13773}

@@

\leavevmode\hypertarget{ref-Poisot2021ImpMam}{}%
Poisot, T., Ouellet, M.-A., Mollentze, N., Farrell, M. J., Becker, D.
J., Albery, G. F., Gibb, R. J., Seifert, S. N., \& Carlson, C. J.
(2021). \emph{Imputing the mammalian virome with linear filtering and
singular value decomposition}. \url{http://arxiv.org/abs/2105.14973}

@@

\leavevmode\hypertarget{ref-Ramasamy2015ComSpe}{}%
Ramasamy, D., \& Madhow, U. (2015). Compressive spectral embedding:
Sidestepping the SVD. In C. Cortes, N. Lawrence, D. Lee, M. Sugiyama, \&
R. Garnett (Eds.), \emph{Advances in neural information processing
systems} (Vol. 28). Curran Associates, Inc.
\url{https://proceedings.neurips.cc/paper/2015/file/4f6ffe13a5d75b2d6a3923922b3922e5-Paper.pdf}

@@

\leavevmode\hypertarget{ref-Stevenson2021AlgRis}{}%
Stevenson, M. T., \& Doleac, J. L. (2021). \emph{Algorithmic Risk
Assessment in the Hands of Humans} {[}SSRN Scholarly Paper{]}.
\url{https://doi.org/10.2139/ssrn.3489440}

\leavevmode\hypertarget{ref-Stier2017IntExp}{}%
Stier, A. C., Samhouri, J. F., Gray, S., Martone, R. G., Mach, M. E.,
Halpern, B. S., Kappel, C. V., Scarborough, C., \& Levin, P. S. (2017).
Integrating Expert Perceptions into Food Web Conservation and
Management. \emph{Conservation Letters}, \emph{10}(1), 67--76.
\url{https://doi.org/10.1111/conl.12245}

@@

\leavevmode\hypertarget{ref-Xu2020UndGra}{}%
Xu, M. (2020). \emph{Understanding graph embedding methods and their
applications}. \url{http://arxiv.org/abs/2012.08019}

@@

\leavevmode\hypertarget{ref-Zhou2020GraNeu}{}%
Zhou, J., Cui, G., Hu, S., Zhang, Z., Yang, C., Liu, Z., Wang, L., Li,
C., \& Sun, M. (2020). Graph neural networks: A review of methods and
applications. \emph{AI Open}, \emph{1}, 57--81.
\url{https://doi.org/10.1016/j.aiopen.2021.01.001}
